\pkg{depmixS4} : An \proglang{R}-package for hidden Markov models

### 2009-07-14 papers/jss/article.tex
Remove pdf (which leads to svn conflicts and cannot be merged).
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/depmix@300 edb9625f-4e0d-4859-8d74-9fd3b1da38cb
Yet another version of the jss paper, now with conclusions as well.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/depmix@301 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

%% an abstract and keywords
\Abstract{ 	
	\pkg{depmixS4} implements a general framework for definining and
	estimating hidden Markov mixture model in the R programming language
	\citep{R2009}.  This includes standard Markov models,
	latent/hidden Markov models, and latent class and finite mixture
	distribution models.  The models can be fitted on mixed
	multivariate data with multinomial and/or gaussian distributions.
	Parameters can be estimated subject to general linear constraints.
	Parameter estimation is done through an EM algorithm or by a
	direct optimization approach using the \pkg{Rdonlp2} optimization
	routine when contraints are imposed on the parameters.  A number
	of illustrative examples are included.
}

@@

Markov and latent Markov models are frequently used in the social
sciences, in different areas and applications.  In psychology, they
are used for modelling learning processes, see \citet{Wickens1982},
for an overview, and \citet{Schmittmann2006} for a recent application.
In economics, latent Markov models are commonly used as regime
switching models, see e.g.\ \citet{Kim1994} and \citet{Ghysels1994}.
Further applications include speech recognition \citep{Rabiner1989},
EEG analysis \citep{Rainer2000}, and genetics \citep{Krogh1998}.  In
those latter areas of application, latent Markov models are usually
referred to as hidden Markov models.

The \pkg{depmixS4} package was motivated by the fact that Markov
models are used commonly in the social sciences, but no comprehensive
package was available for fitting such models.  Common programs for
Markovian models include Panmark \citep{Pol1996}, and for latent class
models Latent Gold \citep{Vermunt2003}.  Those programs are lacking a
number of important features, besides not being freely available.
There are currently some packages in R that handle hidden Markov
models but they lack a number of features that we needed in our
research.  In particular, \pkg{depmixS4} was designed to meet the
following goals:
\begin{enumerate}

@@

Although \pkg{depmixS4} is designed to deal with longitudinal or time
series data, for say $T>100$, it can also handle the limit case with
$T=1$.  In those cases, there are no time dependencies between
observed data, and the model reduces to a finite mixture model, or a
latent class model.  Although there are other specialized packages to
deal with mixture data, one specific feature that we needed ourselves
which is to the best of our knowledge not available in other packages
is the possibility to include covariates on the prior probabilities of
class membership.  In the next section, an outline is provided of the
model and the likelihood equations.

@@

The data considered here, has the general form $O_{1}^{1}, \ldots,
O_{1}^{m}$, $O_{2}^{1}, \ldots, O_{2}^{m}$, \ldots, $O_{T}^{1},
\ldots, O_{T}^{m}$ for an $m$-variate time series of length $T$.  As
an example, consider a time series of responses generated by a single
subject in a reaction time experiment.  The data consists of three
variables, reaction time, accuracy and a covariate which is a pay-off
factor which determines the reward for speed and accuracy.  These
variables are measured on 168, 134 and 137 occasions respectively (in
Figure~\ref{fig:speed} the first part of this series is plotted).

\begin{figure}[htbp]
  \begin{center}
	  \scalebox{1.2}{\includegraphics*{graphs/speed1.pdf}}
	  \caption{Response times, accuracy and pay-off values for
	  the first series of responses in dataset \texttt{speed}.}
	  \label{fig:speed}
  \end{center}
\end{figure}

The latent Markov model is commonly associated with data of this type,
although usually only multinomial variables are considered.  However,
common estimation procedures, such as those implemented in
\citet{Pol1996} are not suitable for long time series due to underflow
problems.  In contrast, the hidden Markov model is typically only used
for `long' univariate time series.  In the next sections, the
likelihood and estimation procedure for the hidden Markov model is
described, given data of the above form.  These models are called
dependent mixture models because one of the authors (Ingmar Visser)
thought it was time for a new name for these models\footnote{Only
later did I find out that \citet{Leroux1992} already coined the term
dependent mixture models in an application with hidden Markov mixtures
of Poisson count data.}

@@

	\item matrices $\mat{A}_t$ of transition probabilities $a_{ij,t}$ for
	the transition from state $S_{i}$ to state $S_{j}$ at time $t$,

	\item a set $\vc{B}_t$ of observation functions $b_j^k(\cdot)$ that
	provide the conditional probabilities of observations $O_{t}^k$ 
	associated with latent state $S_{j}$,

	\item a vector $\pmb{\pi}$ of latent state initial probabilities
	$\pi_{i}$
\end{enumerate}
When transitions are added to the latent class model, it is more
appropriate to refer to the classes as states.  The word class is
rather more associated with a stable trait-like attribute whereas a
state can change over time.

The dependent mixture model is then given by the following equations: 
\begin{align}
	S_{t} = A S_{t-1}, t=2, \ldots, T \\
	O_{t} = f(O_{t}|S_{t}),
\end{align}
where $S_{t}$ is a sequence of hidden states, $A$ is a transition
matrix, $O_{t}$ is an (possibly multivariate) observation and $f$ is a
density function for $O_{t}$ conditional on the hidden state $S_{t}$.
In the example data above, $f$ could be a gaussian distribution
function for the response time variable, and a Bernouilli for the
accuracy data.  In the models we are considering here, both the
transition probabilities $A$ and the initial state probabilities $\pi$
may depend on covariates as well as the response distributions $f$. 
See for example \citet{Fruhwirth2006} for an overview of 
hidden Markov models with extensions.

@@

Parameters are estimated in \pkg{depmixS4} using the EM algorithm or
through the use of a general Newton-Raphson optimizer.

In the EM algorithm, parameters are estimated by iteratively maximising the 
expected joint likelihood of the parameters given the observations and states. 
Let $\greekv{\theta} = (\greekv{\theta}_1, \greekv{\theta}_2,\greekv{\theta}_3)$
be the general parameter vector consisting of three subvectors with parameters 
for the prior model, transition model, and response model respectively. The 
joint log likelihood can be written as
\begin{equation}
\log Pr(O_{1:T}, S_{1:T}|\greekv{\theta}) = \log Pr(S_1|\greekv{\theta}_1) 
+ \sum_{t=2}^{T} \log Pr(S_t|S_{t-1},\greekv{\theta}_2) 
+ \sum_{t=1}^{T} \log Pr(O_t|S_t,\greekv{\theta}_3)
\end{equation}
This likelihood depends on the unobserved states $S_t$. In the Expectation step,
we replace these with their expected values given a set of (initial) parameters 
$\greekv{\theta}' = (\greekv{\theta}'_1, \greekv{\theta}'_2,\greekv{\theta}'_3)$
and observations $O_{1:T}$. The expected log likelihood 
\begin{equation}
Q(\greekv{\theta},\greekv{\theta}') = E_{\greekv{\theta}'} 
(\log Pr(O_{1:T},S_{1:T}|O_{1:T},\greekv{\theta}))
\end{equation}
can be written as
%\begin{equation}
\begin{multline}
\label{eq:Q}
Q(\greekv{\theta},\greekv{\theta}') = 
\sum_{j=1}^n \gamma_1(j) \log Pr(S_1=j|\greekv{\theta}_1) \\ 
+ \sum_{t=2}^T \sum_{j=1}^M \sum_{k=1}^n \xi_t^i(j,k) \log Pr(S_t = k|S_{t-1} 
= j,\greekv{\theta}_2) \\ + \sum_{t=1}^T \sum_{j=1}^n \gamma_t^i(j) 
\ln Pr(O_t|S_t=j,\greekv{\theta}_3),
\end{multline}
%\end{equation}
where the expected values $\xi_t(j,k) =  P(S_t = k, S_{t-1} = j|O_{1:T},\greekv{\theta}')$ and $\gamma_t(j) = P(S_t = j|O_{1:T},\greekv{\theta}')$ can be computed effectively by the Forward-Backward algorithm \citep[see e.g.,][]{Rabiner1989}. The Maximisation step consists of the maximisation of (\ref{eq:Q}) for $\greekv{\theta}$. As the r.h.s. of (\ref{eq:Q}) consists of three separate parts, we can maximise separately for $\greekv{\theta}_1$, $\greekv{\theta}_2$ and $\greekv{\theta}_3$. In common models, maximisation for $\greekv{\theta}_1$ and $\greekv{\theta}_2$ is performed by the \code{nnet.default} routine in \pkg{nnet}, and maximisation for $\greekv{\theta}_3$ by the \code{glm} routine.

@@

The EM algorithm however has some drawbacks.  First, it can be slow to
converge towards the end of optimization (although it is usually
faster than direct optimization at the start, so possibly a
combination of EM and direct optimization is fastest).  Second,
applying constraints to parameters can be problematic; in particular,
EM can lead to wrong parameter estimates when applying constraints.
Hence, in \pkg{depmixS4}, EM is used by default in unconstrained
models, but otherwise, direct optimization is done using \pkg{Rdonlp2}
\cite{Tamura2007,Spellucci2002}, because it handles general linear
(in)equality constraints, and optionally also non-linear constraints.

@@

Two steps are involved in using \pkg{depmixS4} which are illustrated
below with examples:
\begin{enumerate}
	\item model specification with function \code{depmix}

	\item  model fitting with function \code{fit}
\end{enumerate}

@@

\subsection{A simple example}

@@

The so-defined models needs to be \code{fit}ted with the following: 
\begin{CodeChunk}
\begin{CodeInput}
> fm <- fit(mod)
\end{CodeInput}
\end{CodeChunk}

The \code{fit}-function returns an object of class
\code{depmix.fitted} which extends the \code{depmix} class, adding
convergence information (and information about constraints if these
were applied, see below).  The \code{print} method provides summary
information on convergence, the log likelihood and the AIC and BIC
values.  These statistics may be extracted using \code{logLik},
\code{AIC} and \code{BIC}, respectively.

\begin{CodeChunk}
\begin{CodeInput}
> fm
\end{CodeInput}
\begin{CodeOutput}
Convergence info: Log likelihood converged to within tol. 
'log Lik.' -84.34175 (df=7)
AIC:  182.6835 
BIC:  211.275 
\end{CodeOutput}
\end{CodeChunk}

@@

\subsection{Transition parameters}

The transition matrix is parametrized as a list of multinomial
logistic models.  The initial state probabilities are similarly
parametrized as a multinomial logistic model.  Both models use a
baseline category parametrization, meaning that the parameter for the
base category is fixed at zero (see \citet[see][p.\ 267
ff.]{Agresti2002} for multinomial logistic models and various
parameterizations).  The default base category is the first state.
Hence, for example, for a 3-state model, the initial state probability
model would have three parameters of which the first is fixed at zero
and the other two are freely estimated.

Covariates on the transition probabilities can be specified using a
one-sided formula as in the following example:
\begin{CodeChunk}
\begin{CodeInput}
> set.seed(1)
> mod <- depmix(rt~1, data=speed, nstates=2, family=gaussian(), 
  transition=~scale(Pacc), instart=runif(2))
> fm <- fit(mod)
\end{CodeInput}
\begin{CodeOutput}
Initial state probabilties model 
Model of type multinomial, formula: ~1
Coefficients: 
	 [,1]     [,2]
[1,]    0 10.71779
Probalities at zero values of the covariates.
2.214681e-05 0.9999779

Transition model for state (component) 1 
Model of type multinomial, formula: ~scale(Pacc)
Coefficients: 
	 [,1]       [,2]
[1,]    0 -0.9215182
[2,]    0  1.8649734
Probalities at zero values of the covariates.
0.7153513 0.2846487

Transition model for state (component) 2 
Model of type multinomial, formula: ~scale(Pacc)
Coefficients: 
	 [,1]     [,2]
[1,]    0 2.471442
[2,]    0 3.570856
Probalities at zero values of the covariates.
0.07788458 0.9221154

Response model for response 1 
Model of type gaussian, formula: rt ~ 1
Coefficients: 
[1] 5.512179
sd  0.1921098

Response model for response 1 
Model of type gaussian, formula: rt ~ 1
Coefficients: 
[1] 6.3885
sd  0.2402693 
\end{CodeOutput}
\end{CodeChunk}

The summary provides all parameters of the model, also the 
(redundant) zeroes for the base-line category in the multinomial model. 
The summary also prints the transition probabilities at the zero value 
of the covariate. Note that scaling of the covariate is useful in this 
regard as it makes interpretation of these intercept probabilities 
easier.

@@

Multivariate data can be modelled by providing a list of formulae as 
well as a list of family objects for the distributions of the various 
responses. In above examples we have only used the response times 
which were modelled with the gaussian distribution. The accuracy data 
are in the \code{speed} data are modelled with a multinomial by 
specifying the following: 
\begin{verbatim}
set.seed(1)
mod <- depmix(list(rt~1,corr~1), data=speed, nstates=2, 
	family=list(gaussian(),multinomial()), transition=~scale(Pacc), 
	instart=runif(2))
fm <- fit(mod)
\end{verbatim}
which provides the following fitted model parameters (only the 
response parameters are given here): 
\begin{verbatim}
	> summary(fm)
	\ldots 
	Response model(s) for state 1

@@

The \pkg{depmixS4} contains another data set which is used to
illustrate the use of covariates on the prior probabilities of models,
in this case a latent class model. The \code{balance} data consists 
of 4 binary items (correct-incorrect) on a balance scale task 
\citet{Siegler1981}. The data form a subset of the data published in 
\citet{Jansen2002}.

@@

The summary of the \code{fit}ted model gives (only the prior model is 
shown here): 
\begin{verbatim}
> summary(fm)
Mixture probabilities model 
Model of type multinomial, formula: ~age
		[,1]   [,2]
[1,]    0 -2.5182573
[2,]    0  0.5512996
Probalities at zero values of the covariates.
0.9254119 0.07458815 
\end{verbatim}	
Hence at young ages, children have a high probability of incorrect 
responding in class~1, whereas the prior probability for class~2 
increases with age.

@@

Using package \pkg{Rdonlp2}, parameters may be fitted subject to
general linear (in-)equality constraints.  Constraining and fixing
parameters is done using the \code{conpat} argument to the
\code{depmix.fit}-function, which specifies for each parameter in the
model whether it's fixed (0) or free (1 or higher).  Equality
constraints can be imposed by having two parameters have the same
number in the \code{conpat} vector.  When only fixed values are
required the \code{fixed} argument can be used instead of
\code{conpat}, with zeroes for fixed parameters and other values (ones
e.g.) for non-fixed parameters.  Fitting the models subject to these
constraints is handled by the optimization routine \code{donlp2}.

\paragraph{Parameter numbering} When using the \code{conpat} and
\code{fixed} arguments, complete parameter vectors should be supplied,
i.e., these vectors should have length of the number of parameters of
the model, which can be obtained by calling \code{npar(object)}.
Parameters are numbered in the following order:
\begin{enumerate}
	\item  the prior model parameters
	\item  the parameters for the transition models
	\item  the response model parameters per state (and subsequently
	per response in the case of multivariate time series)
\end{enumerate}

@@

When fitting constraints it is useful to have good starting values 
for the parameters and hence we first fit the following model: 
\begin{verbatim}
trst=c(0.9,0.1,0,0,0.1,0.9,0,0)
mod <- depmix(list(rt~1,corr~1),data=speed,transition=~Pacc,
	nstates=2,family=list(gaussian(),multinomial()),
	trstart=trst,inst=c(.999,0.001))
fm <- fit(mod)
\end{verbatim}
After this, we use the fitted values from this model to constrain the 
regression coefficients on the transition matrix:
\begin{verbatim}
# start with fixed and free parameters
conpat <- c(0,1,rep(c(0,1),4),1,1,0,1,1,1,0,1)
# constrain the beta's on the transition parameters to be equal
conpat[6] <- conpat[10] <- 2
fm <- fit(fm,equal=conpat)
\end{verbatim}
Using \code{summary} on the fitted model shows that the regression 
coefficients are now estimated at the same value of 12.66. 
Note that these arguments provide the possibility for arbitrary 
constraints, also between, e.g., a multinomial regression coefficient 
for the transition matrix and the mean of a gaussian response model. 
Whether such constraints make sense is hence the responsibility of 
the user.

@@

Only the constructor function and the fit function are provided here
and the complete code is provided in the helpfile for \code{makeDepmix}. 
The \code{exgaus} example uses the \pkg{gamlss.distr} package for 
computing the \code{dens}ity and for fitting the parameters.

@@

The \code{fit} function defines a trivial \code{gamlss} model with 
only an intercept to be estimated and then sets the fitted parameters 
back into their respective slots in the `exgaus' object. See the help 
for \code{gamlss.distr} for interpretation of these parameters.

@@

\section[Conclusion and future work]{Conclusion \& future work}

What are our next plans?

Adding gradients for speed and computation of the Hessian.

>>>>

%% an abstract and keywords
\Abstract{

	\pkg{depmixS4} implements a general framework for defining and estimating
	 dependent mixture models in the \proglang{R} programming language
	\citep{R2009}. This includes standard Markov models, latent/hidden Markov
	models, and latent class and finite mixture distribution models. The models can
	be fitted on mixed multivariate data with distributions from the \code{glm} 
	family, the logistic multinomial, or the multivariate normal distribution. Other 
	distributions can be added easily, and an example is provided. Parameter 
	estimation is done through an EM algorithm or by a direct optimization approach using
	the\pkg{Rdonlp2} optimization routine when contraints are imposed on the parameters. 
	Parameters can be estimated subject to general linear constraints.  
}

@@

Markov and latent Markov models are frequently used in the social
sciences, in different areas and applications.  In psychology, they
are used for modelling learning processes, see \citet{Wickens1982},
for an overview, and \citet{Schmittmann2006} for a recent application.
In economics, latent Markov models are commonly used as regime
switching models, see e.g.\ \citet{Kim1994} and \citet{Ghysels1994}.
Further applications include speech recognition \citep{Rabiner1989},
EEG analysis \citep{Rainer2000}, and genetics \citep{Krogh1998}.  In
those latter areas of application, latent Markov models are usually
referred to as hidden Markov models. See e.g.\ \citet[][chapter~1]{Cappe2005}
for an overview of different applications.

The \pkg{depmixS4} package was motivated by the fact that Markov
models are used commonly in the social sciences, but no comprehensive
package was available for fitting such models.  Common programs for
Markovian models include Panmark \citep{Pol1996}, and for latent class
models Latent Gold \citep{Vermunt2003}.  Those programs are lacking a
number of important features, besides not being freely available.
There are currently some packages in R that handle hidden Markov
models but they lack a number of features that we needed in our
research.  In particular, \pkg{depmixS4} was designed to meet the
following goals:

\begin{enumerate}

@@

Although \pkg{depmixS4} is designed to deal with longitudinal or time
series data, for say $T>100$, it can also handle the limit case with
$T=1$.  In those cases, there are no time dependencies between
observed data, and the model reduces to a finite mixture model, or a
latent class model.  Although there are other specialized packages to
deal with mixture data, one specific feature that we needed ourselves
which is to the best of our knowledge not available in other packages
is the possibility to include covariates on the prior probabilities of
class membership and the transition probabilities. In the next section, 
an outline is provided of the model and the likelihood equations.

@@

The data considered here, has the general form $O_{1}^{1}, \ldots,
O_{1}^{m}$, $O_{2}^{1}, \ldots, O_{2}^{m}$, \ldots, $O_{T}^{1},
\ldots, O_{T}^{m}$ for an $m$-variate time series of length $T$.  As
an example, consider a time series of responses generated by a single
participant in a response time experiment.  The data consists of three
variables, response time, accuracy and a covariate which is a pay-off
determining the relative reward for speeded and/or accuracy responding. These
variables are measured on 168, 134 and 137 occasions respectively (in
Figure~\ref{fig:speed} the first part of this series is plotted). These data
are more fully described in \citet{Dutilh2009}.

\begin{figure}[htbp]
  \begin{center}
	  \scalebox{1.2}{\includegraphics*{graphs/speed1.pdf}}
	  \caption{Response times (rt), accuracy (corr) and pay-off values (Pacc) for
	  the first series of responses in dataset \code{speed}.}
	  \label{fig:speed}
  \end{center}
\end{figure}

The latent Markov model is commonly associated with data of this type,
although usually only multinomial response variables are considered.  However,
common estimation procedures, such as those implemented in
\citet{Pol1996} are not suitable for long time series due to underflow
problems.  In contrast, the hidden Markov model is typically only used
for `long' univariate time series 
\cite[see e.g.][, chapter~1 for an overview of examples]{Cappe2005}.  
In the next sections, the
likelihood and estimation procedures for the dependent mixture model is
described for data of the above form.  These models are called
dependent mixture models because one of the authors (Ingmar Visser)
thought it was time for a new name for these models\footnote{Only
later did I find out that \citet{Leroux1992} already coined the term
dependent mixture models in an application with hidden Markov mixtures
of Poisson count data.}.

@@

	\item matrices $\mat{A}$ of transition probabilities $a_{ij}$ for
	the transition from state $S_{i}$ to state $S_{j}$,

	\item a set $\vc{B}$ of observation density functions $b_j^k(\cdot)$ that
	provide the conditional densities of observations $O_{t}^k$ 
	associated with latent state $S_{j}$,

	\item a vector $\pmb{\pi}$ of latent state initial probabilities
	$\pi_{i}$
\end{enumerate}

The dependent mixture model is defined by the following equations: 
\begin{align}
	S_{t} &= A S_{t-1}, t=2, \ldots, T \\
	O_{t} &= b(O_{t}|S_{t}), 
\end{align}
where $S_{t}$ is a sequence of hidden states, $A$ is a transition
matrix, $O_{t}$ is an (possibly multivariate) observation and $b$ is a
density function for $O_{t}$ conditional on the hidden state $S_{t}$.
In the example data above, $b$ could be a gaussian distribution
function for the response time variable, and a Bernouilli for the
accuracy data.  In the models we are considering here, both the
transition probabilities $\mat{A}$ and the initial state probabilities $\pi$
may depend on covariates as well as the response distributions $\vc{B}$. 
See for example \citet{Fruhwirth2006} for an overview of 
hidden Markov models with extensions.

@@

Parameters are estimated in \pkg{depmixS4} using the EM algorithm or
through the use of a general Newton-Raphson optimizer.  In the EM algorithm, 
parameters are estimated by iteratively maximising the 
expected joint likelihood of the parameters given the observations and states. 
Let $\greekv{\theta} = (\greekv{\theta}_1, \greekv{\theta}_2,\greekv{\theta}_3)$
be the general parameter vector consisting of three subvectors with parameters 
for the prior model, transition model, and response model respectively. The 
joint log likelihood can be written as
\begin{equation}
\log \Prob(O_{1:T}, S_{1:T}|\greekv{\theta}) = \log \Prob(S_1|\greekv{\theta}_1) 
+ \sum_{t=2}^{T} \log \Prob(S_t|S_{t-1},\greekv{\theta}_2) 
+ \sum_{t=1}^{T} \log \Prob(O_t|S_t,\greekv{\theta}_3)
\end{equation}
This likelihood depends on the unobserved states $S_t$. In the Expectation step,
we replace these with their expected values given a set of (initial) parameters 
$\greekv{\theta}' = (\greekv{\theta}'_1, \greekv{\theta}'_2,\greekv{\theta}'_3)$
and observations $O_{1:T}$. The expected log likelihood 
\begin{equation}
Q(\greekv{\theta},\greekv{\theta}') = E_{\greekv{\theta}'} 
(\log \Prob(O_{1:T},S_{1:T}|O_{1:T},\greekv{\theta}))
\end{equation}
can be written as
%\begin{equation}
\begin{multline}
\label{eq:Q}
Q(\greekv{\theta},\greekv{\theta}') = 
\sum_{j=1}^n \gamma_1(j) \log \Prob(S_1=j|\greekv{\theta}_1) \\ 
+ \sum_{t=2}^T \sum_{j=1}^M \sum_{k=1}^n \xi_t^i(j,k) \log \Prob(S_t = k|S_{t-1} 
= j,\greekv{\theta}_2) \\ + \sum_{t=1}^T \sum_{j=1}^n \gamma_t^i(j) 
\ln \Prob(O_t|S_t=j,\greekv{\theta}_3),
\end{multline}
%\end{equation}
where the expected values $\xi_t(j,k) =  P(S_t = k, S_{t-1} = j|O_{1:T},\greekv{\theta}')$ and $\gamma_t(j) = P(S_t = j|O_{1:T},\greekv{\theta}')$ can be computed effectively by the Forward-Backward algorithm \citep[see e.g.,][]{Rabiner1989}. The Maximisation step consists of the maximisation of (\ref{eq:Q}) for $\greekv{\theta}$. As the r.h.s. of (\ref{eq:Q}) consists of three separate parts, we can maximise separately for $\greekv{\theta}_1$, $\greekv{\theta}_2$ and $\greekv{\theta}_3$. In common models, maximisation for $\greekv{\theta}_1$ and $\greekv{\theta}_2$ is performed by the \code{nnet.default} routine in \pkg{nnet}, and maximisation for $\greekv{\theta}_3$ by the \code{glm} routine.

@@

The EM algorithm however has some drawbacks.  First, it can be slow to
converge towards the end of optimization.  Second,
applying constraints to parameters can be problematic; in particular,
EM can lead to wrong parameter estimates when applying constraints.
Hence, in \pkg{depmixS4}, EM is used by default in unconstrained
models, but otherwise, direct optimization is done using \pkg{Rdonlp2}
\cite{Tamura2009,Spellucci2002}, because it handles general linear
(in)equality constraints, and optionally also non-linear constraints.

@@

Two steps are involved in using \pkg{depmixS4} which are illustrated
below with examples:
\begin{enumerate}
	\item model specification with function \code{depmix} (or with \code{mix}
	for latent class and finite mixture models, see example below on adding
	covariates to prior probabilities)

	\item  model fitting with function \code{fit}
\end{enumerate}
We have separated the stages of model specification and model fitting because
fitting large models can be fairly time-consuming and it is hence useful to be
able to check the model specification before actually fitting the model.

@@

\subsection{A simple model}

@@

The so-defined models needs to be \code{fit}ted with the following
line of code:
\begin{CodeChunk}
\begin{CodeInput}
> fm <- fit(mod)
\end{CodeInput}
\end{CodeChunk}

The \code{fit}-function returns an object of class
\code{depmix.fitted} which extends the \code{depmix} class, adding
convergence information (and information about constraints if these
were applied, see below).  The \code{print} method provides summary
information on convergence, the log likelihood and the AIC and BIC
values:
\begin{CodeChunk}
\begin{CodeInput}
> fm
\end{CodeInput}
\begin{CodeOutput}
Convergence info: Log likelihood converged to within tol. 
'log Lik.' -84.34175 (df=7)
AIC:  182.6835 
BIC:  211.275 
\end{CodeOutput}
\end{CodeChunk}
These statistics may be extracted using \code{logLik},
\code{AIC} and \code{BIC}, respectively.

@@

\subsection{Covariates on transition parameters}

The transition probabilities and the initial state probabilities
are parameterized using the multinomial
logistic model.   In particular, each row of the transition matrix is
parameterized by a baseline category logistic multinomial, 
meaning that the parameter for the
base category is fixed at zero (see \citet[see][p.\ 267
ff.]{Agresti2002} for multinomial logistic models and various
parameterizations). See also \citet{Chung2007} for similar models, latent transition models using logistic regression on the transition parameters. They fit such models on repeated measurement
data ($T=2$) using Bayesian methods.  The default baseline category is the first state.
Hence, for example, for a 3-state model, the initial state probability
model would have three parameters of which the first is fixed at zero
and the other two are freely estimated.

Covariates on the transition probabilities can be specified using a
one-sided formula as in the following example:
\begin{CodeChunk}
\begin{CodeInput}
set.seed(1)
mod <- depmix(rt~1, data=speed, nstates=2, family=gaussian(), 
  transition=~scale(Pacc), instart=runif(2))
fm <- fit(mod)
\end{CodeInput}
\end{CodeChunk}

Applying \code{summary} to the \code{fit}ted object gives (only transition models
printed here): 
\begin{CodeChunk}
\begin{CodeOutput}
...
Transition model for state (component) 1 
Model of type multinomial, formula: ~scale(Pacc)
Coefficients: 
	 [,1]       [,2]
[1,]    0 -0.9215182
[2,]    0  1.8649734
Probalities at zero values of the covariates.
0.7153513 0.2846487

Transition model for state (component) 2 
Model of type multinomial, formula: ~scale(Pacc)
Coefficients: 
	 [,1]     [,2]
[1,]    0 2.471442
[2,]    0 3.570856
Probalities at zero values of the covariates.
0.07788458 0.9221154
...
\end{CodeOutput}
\end{CodeChunk}
The summary provides all parameters of the model, also the 
(redundant) zeroes for the base-line category in the multinomial model. 
The summary also prints the transition probabilities
at the zero value of the covariate. Note that scaling of the covariate 
is useful in this regard as it makes interpretation of these intercept probabilities 
easier.

@@

Multivariate data can be modelled by providing a list of formulae as 
well as a list of family objects for the distributions of the various 
responses. In above examples we have only used the response times 
which were modelled with the gaussian distribution. The accuracy data 
are in the \code{speed} data are modelled with a multinomial by 
specifying the following: 
\begin{verbatim}
set.seed(1)
mod <- depmix(list(rt~1,corr~1), data=speed, nstates=2, 
	family=list(gaussian(),multinomial()),
	 transition=~scale(Pacc),instart=runif(2))
fm <- fit(mod)
\end{verbatim}
which provides the following fitted model parameters (only the 
response parameters are given here): 
\begin{verbatim}
	> summary(fm)
	...
	Response model(s) for state 1

@@

To illustrate the use of covariates on the prior probabilities we have included
another data set with \pkg{depmixS4}. The \code{balance} data consists 
of 4 binary items (correct-incorrect) on a balance scale task 
\citet{Siegler1981}. The data form a subset of the data published in 
\citet{Jansen2002}.

@@

The summary of the \code{fit}ted model gives (only the prior model is 
shown here): 
\begin{verbatim}
> summary(fm)
Mixture probabilities model 
Model of type multinomial, formula: ~age
		[,1]   [,2]
[1,]    0 -2.5182573
[2,]    0  0.5512996
Probalities at zero values of the covariates.
0.9254119 0.07458815 
...
\end{verbatim}	
Hence at young ages, children have a high probability of incorrect 
responding in class~1, whereas the prior probability for class~2 
increases with age.

@@

Using package \pkg{Rdonlp2} by \citet{Tamura2009}, parameters may be fitted subject to
general linear (in-)equality constraints.  Constraining and fixing
parameters is done using the \code{conpat} argument to the
\code{fit}-function, which specifies for each parameter in the
model whether it's fixed (0) or free (1 or higher).  Equality
constraints can be imposed by having two parameters have the same
number in the \code{conpat} vector.  When only fixed values are
required, the \code{fixed} argument can be used instead of
\code{conpat}, with zeroes for fixed parameters and other values (ones
e.g.) for non-fixed parameters.  Fitting the models subject to these
constraints is handled by the optimization routine \code{donlp2}.
To be able to construct the \code{conpat} and/or \code{fixed} vectors 
one needs the correct ordering of parameters which is briefly discussed 
next before proceeding with an example.

\paragraph{Parameter numbering} When using the \code{conpat} and
\code{fixed} arguments, complete parameter vectors should be supplied,
i.e., these vectors should have length of the number of parameters of
the model, which can be obtained by calling \code{npar(object)}. Note that
this is not the same as the degrees of freedom used e.g.\ in the \code{logLik}
function because \code{npar} also counts the baseline category zeroes
from the multinomial logistic models. 
Parameters are numbered in the following order:
\begin{enumerate}
	\item  the prior model parameters
	\item  the parameters for the transition models
	\item  the response model parameters per state (and subsequently
	per response in the case of multivariate time series)
\end{enumerate}

@@

When fitting constraints it is useful to have good starting values 
for the parameters and hence we first fit the following model without
constraints:
\begin{verbatim}
trst=c(0.9,0.1,0,0,0.1,0.9,0,0)
mod <- depmix(list(rt~1,corr~1),data=speed,transition=~Pacc,
	nstates=2,family=list(gaussian(),multinomial()),
	trstart=trst,inst=c(.999,0.001))
fm1 <- fit(mod)
\end{verbatim}
After this, we use the fitted values from this model to constrain the 
regression coefficients on the transition matrix (parameters numbers~6 and~10):
\begin{verbatim}
# start with fixed and free parameters
conpat <- c(0,1,rep(c(0,1),4),1,1,0,1,1,1,0,1)
# constrain the beta's on the transition parameters to be equal
conpat[6] <- conpat[10] <- 2
fm2 <- fit(fm,equal=conpat)
\end{verbatim}
Using \code{summary} on the fitted model shows that the regression 
coefficients are now estimated at the same value of 12.66. The function
\code{llratio} computes the likelihood ratio $\chi^2$-statistic and the
associated $p$-value with appropriate degrees of freedom for testing the
tenability of constraints \citep{Dannemann2007}. Note that these arguments 
provide the possibility for arbitrary 
constraints, also between, e.g., a multinomial regression coefficient 
for the transition matrix and the mean of a gaussian response model. 
Whether such constraints make sense is hence the responsibility of 
the user.

@@

Only the constructor function and the fit function are provided here
and the complete code is provided in the helpfile for \code{makeDepmix}. 
The \code{exgaus} example uses the \pkg{gamlss} and \pkg{gamlss.distr} packages \cite{Stasinopoulos2009a,Stasinopoulos2009b} for 
computing the \code{dens}ity and for \code{fit}ting the parameters.

@@

The \code{fit} function defines a \code{gamlss} model with 
only an intercept to be estimated and then sets the fitted parameters 
back into their respective slots in the `exgaus' object. See the help 
for \code{gamlss.distr} for interpretation of these parameters.

@@

\section[Conclusions and future work]{Conclusions \& future work}

\pkg{depmixS4} provides a flexible framework for fitting dependent mixture models
for a large variety of response distributions. It can also fit latent class regression
and finite mixture models, although it should be noted that more specialized packages 
are available for doing so, e.g.\ \pkg{FlexMix} \cite{Leisch2004}. The package is intended
for modeling of (individual) time series data with the aim of characterizing the transition 
processes underlying the data. The possibility to use covariates on the transition matrix
greatly enhances the flexibility of the model. The EM algorithm uses a very general 
interface that allows easy addition of new response models.

We are currently working on implementing the gradients for response and transition models 
with two goals in mind. First, to speed up (constrained) parameter optimization using 
\pkg{Rdonlp2}. Second, analytic gradients are useful in computing the Hessian of the
estimated parameters so as to arrive at standard errors for those. We are also planning to
implement goodness-of-fit statistics \citep{Titman2008}.
